---
title: "Indian Housing (II)"
subtitle: "Philosophy, Politics, and Economics"
author: "Miguel A. Arranz"
date: today
date-format: long
toc: true
format: 
    html:
        code-fold: true
        code-summary: "Show the code"
---



# Data and Packages

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(modelsummary)
library(jtools)
library(haven)
```

```{r message=FALSE, warning=FALSE}
library(here)
df <- read_csv(here("Data", "IndianHousingClean.csv"))
df$beds <- df$`No. of Bedrooms`
df$lprice <- log(df$Price)
df$larea <- log(df$Area)
```


# Our First Model

$$
\log Price  = \beta_0 + \beta_1 \log Area + \beta_2 SwimmingPool + \beta_3 AC + \beta_4 beds +  u
$${#eq-1}

In @eq-1 all the coeffcients are the same for all the cities.


```{r}
model1 <- lm(lprice ~ larea + SwimmingPool + AC + beds , data=df)
summ(model1)
```


# Different Intercepts

$$
\begin{aligned}
\log Price & = \beta_0 + \beta_1 \log Area + \beta_2 SwimmingPool + \beta_3 AC + \beta_4 beds\\
 & + \beta_5 Mumbai + \beta_6 Delhi + \beta_7 Chennai+ \beta_8 Hyderabad + \beta_9 Kolkata +  u
\end{aligned}
$${#eq-2}

In @eq-2 the intercepts are allowed to be different for each city, but the rest of the parameters must be equal.


```{r}
model2 <- lm(lprice ~ larea + SwimmingPool + AC + beds +  City_Mumbai + City_Delhi +
  City_Chennai + City_Hyderabad + City_Kolkata , data=df)
summ(model2)
```
Let's assume that $\beta_9=0$. What is the meaning?

# Different Intercepts and Slopes of log Area

$$
\begin{aligned}
\log Price & = \beta_0 + \beta_1 \log Area + \beta_2 SwimmingPool + \beta_3 AC + \beta_4 beds\\
 & + \beta_5 Mumbai + \beta_6 Delhi + \beta_7 Chennai+ \beta_8 Hyderabad + \beta_9 Kolkata \\
 & + \beta_{10} \log Area*Mumbai + \beta_{11} \log Area*Delhi + \beta_{12} \log Area*Chennai \\ 
 & + 
 \beta_{13} \log Area* Hyderabad + \beta_{14} \log Area*Kolkata  +  u
\end{aligned}
$${#eq-3}


```{r}
model3 <- lm(lprice ~ (1 + larea)*(City_Mumbai + City_Delhi +
  City_Chennai + City_Hyderabad + City_Kolkata) + SwimmingPool + AC + beds  , data=df)
summ(model3)
```

# Monster Model


$$
\begin{aligned}
\log Price & = \beta_0 + \beta_1 \log Area + \beta_2 SwimmingPool + \beta_3 AC + \beta_4 beds\\
 & + \beta_5 Mumbai + \beta_6 Delhi + \beta_7 Chennai+ \beta_8 Hyderabad + \beta_9 Kolkata \\
 & + \beta_{10} \log Area*Mumbai + \beta_{11} \log Area*Delhi + \beta_{12} \log Area*Chennai \\ 
 & + 
 \beta_{13} \log Area* Hyderabad + \beta_{14} \log Area*Kolkata  \\
  & + \beta_{15} SwimmingPool*Mumbai + \beta_{16} SwimmingPool*Delhi + \beta_{17} SwimmingPool*Chennai \\ 
 & + 
 \beta_{18} SwimmingPool* Hyderabad + \beta_{19} SwimmingPool*Kolkata  \\
  & + \beta_{20} AC*Mumbai + \beta_{21} AC*Delhi + \beta_{22} AC*Chennai \\ 
 & + 
 \beta_{23} AC* Hyderabad + \beta_{24} AC*Kolkata  \\
 & + \beta_{25} beds*Mumbai + \beta_{26} beds*Delhi + \beta_{27} beds*Chennai \\ 
 & + 
 \beta_{28} beds* Hyderabad + \beta_{29} beds*Kolkata
 +  u
\end{aligned}
$${#eq-4}




```{r}
model4 <- lm(lprice ~ (1 + larea + SwimmingPool + AC+ beds)*(City_Mumbai + City_Delhi +
  City_Chennai + City_Hyderabad + City_Kolkata)  , data=df)
summ(model4)
```

# Partitioned Regression

Remember @eq-1

$$
\log Price  = \beta_0 + \beta_1 \log Area + \beta_2 SwimmingPool + \beta_3 AC + \beta_4 beds +  u
$$


```{r}
summ(model1)
```
Let's calculate the estimated value of $\beta_1$ using the partitioned regression thorem

1. Estimate the Model

$$
\log Price  = \delta_0 + \delta_1 SwimmingPool + \delta_2 AC + \delta_3 beds +  \nu
$${#eq-5}
and save the residuals to $\hat{\nu}$

2. Estimate the model
$$
\log Area  = \gamma_0 + \gamma_1 SwimmingPool + \gamma_2 AC + \gamma_3 beds +  \eta
$${#eq-6}
and save the residuals to $\hat{\eta}$

3. Run the regression of $\hat{nu}$ on $\hat{eta}$. The slope will be $\hat{\beta}_1$

Estimate @eq-5 and save residuals

```{r}
model5 <- lm(lprice ~  SwimmingPool + AC + beds , data=df)
summ(model5)
nuhat <- model5$residuals
```

Estimate @eq-6 and save residuals

```{r}
model6 <- lm(larea ~  SwimmingPool + AC + beds , data=df)
summ(model6)
etahat <- model6$residuals
```

Run simple linear regression 

```{r}
model7 <- lm(nuhat ~  etahat)
summ(model7)
```

# Quadratic Effects of the Area


$$
\log Price  = \beta_0 + \beta_1 \log Area + \beta_2 (\log Area)^2 + \beta_3 SwimmingPool + \beta_4 AC + \beta_5 beds +  u
$${#eq-8}

Estimate @eq-8


```{r}
model8 <- lm(lprice ~ larea + I(larea^2) + SwimmingPool + AC + beds , data=df)
summ(model8)
```

Note that the marginal effect of $\log Area$ is initially negative. What is the turning point?

$$
\begin{aligned}
\dfrac{\partial \log Price}{\log Area} & = \beta_1 + 2 \beta_2 \log Area \\
x^\star & = \dfrac{- \beta_1}{2 \beta_2}
\end{aligned}
$$

```{r}
betas <- model8$coefficients
names(betas)
```

```{r}
num <- -betas["larea"]
den <- 2*betas["I(larea^2)"]
xstar <- num/den 
cat("Turning Point (log Area): ", xstar, "\n")
cat("Turning Point (Area): ", exp(xstar), "\n")
```




# Linear Probability Model



$$
\begin{aligned}
SwimmingPool & = \beta_0 + \beta_1 \log Area + \beta_2 \log Price + \beta_3 AC + \beta_4 beds +  u
\end{aligned}
$${#eq-9}

In @eq-9 we are modeling the probaility that a house has swimming pool conditional to the values of the regressors. This is what we call the **linear probability model**. 


```{r}
model9 <- lm(SwimmingPool ~ larea + lprice + AC + beds  , data=df)
summ(model9)
```


## Second LPM Model


$$
\begin{aligned}
SwimmingPool & = \beta_0 + \beta_1 \log Area + \beta_2 \log Price + \beta_3 AC + \beta_4 beds\\
 & + \beta_5 Mumbai + \beta_6 Delhi + \beta_7 Chennai+ \beta_8 Hyderabad + \beta_9 Kolkata +  u
\end{aligned}
$${#eq-10}

In @eq-10 the intercepts are allowed to be different for each city, but the rest of the parameters must be equal.


```{r}
model10 <- lm(SwimmingPool ~ larea + lprice + AC + beds +  City_Mumbai + City_Delhi +
  City_Chennai + City_Hyderabad + City_Kolkata , data=df)
summ(model10)
```
