---
title: "The Sample Mean"
subtitle: "Philosophy, Politics, and Economics"
author: "Miguel A. Arranz"
date: today
date-format: long
format: 
    html:
        code-fold: true
        code-summary: "Show the code"
---


# Data

```{r warning=FALSE, message=FALSE}
library(foreign)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(rmarkdown)
library(kableExtra)
library(modelsummary)
library(plotly)
library(haven)
```

```{r}
library(here)
df7 <- read_dta(here("Data","AED_COINTOSSMEANS.dta"))
df <- read_dta(here("Data","AED_CENSUSAGEMEANS.dta"))
df2 <- read_dta(here("Data", "AED_EARNINGS.dta"))
df3 <- read_dta(here("Data","AED_EARNINGSMALE.dta"))
df4 <- rbind(df2, df3)
```

```{r}
df4$educ <- cut(df4$education, breaks = c(-Inf, 12, 16, Inf), labels = c("HighSchool", "College", "PostGraduate"))
```

```{r}
datasummary_skim(df4)
```

# Categorical Variables

```{r}
df4$gender_categorical <- factor(df4$gender, levels = c(0, 1), labels = c("Female", "Male"))
```

```{r}
datasummary_skim(df4, type="categorical")
```

# Introduction

Consider **statistical inference**

- extrapolating from sample to population
- form sample mean $\bar{x}$ to population mean $\mu$

Basic idea is that the sample values $x_1, x_2, \ldots, x_n$ are **realizations of random variables** $X_1, X_2, \ldots , X_n$ 

$\bar{x}$ is a **random variable**

# Random Variables 

A **random variable **is a variable whose value is determined by the
outcome of an experiment.

An **experiment** is an operation whose outcome cannot be predicted
with certainty.



Example: the experiment is tossing a coin and the random variable
takes value 1 if heads and 0 if tails.



Example: the experiment is randomly selecting a person from the
population and the associated random variable takes value equal to
their annual earnings.

**Standard notation**

- $X$ (or $Y$ or $Z$) denotes a random variable
- $x$ (or $y$ or $z$) denotes the values taken by $X$ (or $Y$ or $Z$).


## Example: Coin toss

Simplest case is a random variable that takes only 2 possible values

Consider toss of fair coin with $X = 1$ if heads and $X = 0$ if tails.

$$X = 
\begin{cases}
 0   &\text{with probability } 0.5\\
 1    &\text{with probability } 0.5\\
\end{cases}$$

# Mean of Random Variable

Mean of $X$, denoted as $\mu$ or $\mu_X$

- is the probability-weighted average of all possible values of $X$ in the
population.
- $\mu$ is also denoted as $\mathbb{E}[X]$ (**expected value** of $X$)
- the long-run average value expected if we draw a value of $X$ at
random, draw a second value of $X$ at random, and so on, and then
obtain the average of these values.



$$
\mu = \mathbb{E}[X] = \sum_x x \times \Pr(X=x)
$$

In this case

$$
\begin{aligned}
\mu & = 0 \times \Pr(X=0) + 1 \times \Pr(X=1) \\
& = 0 \times 0.5 + 1 \times 0.5 \\
& = 0.5 
\end{aligned}
$$

What is the expected value if the probability of getting tails is 0.6?

# Variance and Standard Deviation

**Variance** $\sigma^2$

- measures the variability in $X$ around $\mu$.
- equals the expected value of $(X-\mu)^2$
- probability weighted average
$$
\sigma^2 = \sum_x (x-\mu)^2 \times \Pr(X=x)
$$



**Example**

$$
\begin{aligned}
\sigma^2 & = (0 - \mu)^2 \times \Pr(X=0) + (1 - \mu)^2 \times \Pr(X=1) \\
& =  (0 - 0.5)^2 \times \Pr(X=0) + (1 - 0.5)^2 \times \Pr(X=1) \\
& = 0.25
\end{aligned}
$$

**Population standard deviation** $\sigma$ is the square root of the variance


# Random Samples

- A sample of size $n$ takes values denoted as $x_1, x_2, \dots x_n$ 
- These values are realizations or outcomes of the random variables $X_1, X_2, \ldots, X_n$
- Example: four xonsecutive coin tosses with results tails, heads, heads, and heads.
 * $x_1=0, x_2=1, x_3=1, x_4=1$



# Sample Mean is a Random Variable

**Sample** of sie $n$ has observed values $x_1, x_2, \ldots x_n$.
- These are realizations of the random variable

**Sample mean** is the average

$$\bar{x} = \frac{1}{n} \sum_{i=1}^n x_i$$


This is a realization of the **random variable** 

$$\bar{X} = \frac{1}{n} \sum_{i=1}^n X_i$$


# Sample Generated by Experiment: Coin tosses

We consider a simple experiment that generates many samples

- hence many sample means $\bar{x}$
- then summarize the resulting distribution of the many $\bar{x}$.

**Population**: Outcomes from experiment of tossing a coin

- $X = 1$ if heads and $X = 0$ if tails
- Population mean $\mu = \mathbb{E}[X] = 0.5$ and standard deviation $\sigma = 0.5$.

Sample: n = 30

- random sample of size 30 from 30 coin tosses
- there are 10 heads and 20 tails, so x$\bar{x}$ = 10/30 = 0.333
I histogram of this single sample is given in next slide.

# Tossing the Coin

```{r}

set.seed(10101)
u <- runif(30, min=0, max=1)
x <- ifelse(u > 0.5, 1, 0)

# Summary of the first dataset
#summary(cbind(x, u))
#table(x)

hist(x, xlab = "Heads=1, Tails=0")



```

Randomly draw 400 diﾂ､erent samples, each of size 30
- then $\bar{x}_1 = .333, \bar{x}_2 = .500, \bar{x}_3 = 533, \dots$

Histogram (plus kernel density estimate) for the 400 means from the
400 samples of size 30 is given in next slide.

- roughly centered on the population 
  * the average of the 400 means is 0.499, close to $\mu$ = 0.5.



- much less variability in these 400 means than in the original population
  * the standard deviation of the 400 means is 0.086
  * much less than the population standard deviation of $\sigma$ = 0.5

- the density estimate is roughly that of the normal.



```{r}
hist(df7$xbar, xlab = "Sample mean of each of 400 samples", freq = FALSE)
x <- seq(0, 1, by=0.02)
curve(dnorm(x, mean(df7$xbar), sd(df7$xbar)), add=TRUE)
```

# Properties of the Sample Mean

The properties of $\bar{X}$ depend on the properties of $X_1, X_2, \dots, X_n$

- such as the means and variances of $X_1, X_2, \dots, X_n$
- and whether their values depend in part on other values.



Simplest and standard set of
assumptions in introductory statistics

- $X_1, X_2, \dots, X_n$ have common mean, $\mu$, and common variance, $\sigma^2$
- $X_1, X_2, \dots, X_n$ are statistically independent

We will have to relax these assumptions later on.

# Population Assumptions

**Population**

set of all observations (or experimental outcomes).

**Sample**

subset selected from the population.

Properties of $\bar{x}$ depend on the random variable $\bar{X}$
I
hence on assumptions about process generating $X_1, X_2, \ldots , X_n$



We assume a **simple random sample** where

- $X_i$, share a **common  mean**, $\mu$, $\mu=\mathbb{E}[X_i] = \mu$, $\forall i$
- $X_i$, share a **common  variance**, $\sigma$, $\mu=\mathbb{V}[X_i] = \sigmaﾂｲ$, $\forall i$
- $X_i$ is **statistically indepemdent** of $X_j$ $\forall i \neq j$

Shorthand notation

$$
X_i \overset{iid}{\sim} (\mu, \sigmaﾂｲ)
$$

# Mean and Variance of the Sample Mean

Consider $X_i \overset{iid}{\sim} (\mu, \sigmaﾂｲ)$

$$
\bar{X} = \frac{1}{n} \sum_{i=1}^n X_i
$$

- $\mu_{\bar{X}} = \mathbb{E}[\bar{X}] = \mu$
- $\sigmaﾂｲ_{\bar{X}} = \mathbb{V}[\bar{X}] = \frac{\sigmaﾂｲ}{n}$



Sample mean is less variable than the underlying data

$$
\sigmaﾂｲ_{\bar{X}} < \sigma^2
$$

Sample mean is close to $\mu$ as $n \rightarrow \infty$


since $\mathbb{E}[\bar{X}] = \mu$ and $\sigmaﾂｲ_{\bar{X}} \rightarrow 0$ as $n \rightarrow \infty$.


# Normal Distribution and the Central Limit Theorem

We have shown that $\bar{X} \sim \big(\mu , \frac{\sigmaﾂｲ}{n} \big)$

Subtracting the mean and dividing by the standard
deviation yields a random variable with mean 0 and variance 1.


$$
Z = \frac{\bar{X} - \mu}{\sqrt{\sigmaﾂｲ/n}} \sim (0,1)
$$

The central limit theorem (a remarkable result) proves normality as
the sample size gets large

$$
Z \sim N(0,1) \quad \text{ as } n \rightarrow \infty
$$

The CLT works under the current and also weaker assumptions. 



For $n \rightarrow \infty$:

$$
 Z = \frac{\bar{X} - \mu}{\sqrt{\sigmaﾂｲ/n}} \sim N (0,1)
$$


As $n \rightarrow \infty$:

$$
\bar{X} \overset{a}{\sim} N(\mu, \sigmaﾂｲ/n)
$$

$\sigmaﾂｲ$ is unknown 

- we will have to get an estimate
- replace $\sigmaﾂｲ$ by its estimate $sﾂｲ$
- where $s$ is the sample standard deviation of $X$.


# Estimation of the Sample Mean

Desire a good point estimate of population mean $\mu$

Why use $\bar{x}$ rather than some other estimate?


A desirable estimator of $\mu$ has distribution

- centered on $\mu$
- with as little variability around $\mu$ as possible.

# Parameter, Estimator and Estimate

A **parameter** is a constant that determines in part the distribution of
$X$.

An **estimator** is a method for estimating a parameter.

An **estimate** is the particular value of the estimator obtained from
the sample.

For estimation of the mean of $X$ using the sample mean

- the parameter is $\mu$
- the estimator is the random variable $\bar{X}$
- the estimate is the sample value $\bar{x}$.

# Unbiased Estimators

An **unbiased estimator** of a population parameter
has expected value that equals the population parameter.

The sample mean is unbiased for $\mu$

$$
\mathbb{E}[\bar{X}] = \mu
$$



# Minimum Variance Estimators

Other estimators may also be unbiased and consistent for $\mu$

- e.g. sample median in the case where $X$ is symmetrically distributed
- discriminate between such estimators using their variance.

A best estimator or efficient estimator

has minimum variance among the class of consistent estimators (or of
unbiased estimators).


Under the three assumptions, the sample mean has variance $\sigmaﾂｲ/n$

- for $X$ that is normal, Bernoulli, binomial or Poisson no other unbiased
estimator has lower variance
- for $X$ with other distributions the sample mean is often close to having
the lowest variance
- generally the sample mean is used to estimate $\mu$.


# Consistent Estimators

**Consistency** is a more advanced concept that considers behavior as
the sample size goes to infinity.

A consistent estimator of a population parameter

is one that is almost certainly arbitrarily close to the population
parameter as the sample size gets very large.



A sufficient condition for consistency is

- any bias disappears as the sample size gets very large
- the variance goes to zero as the sample size gets very large


The sample mean is consistent under the assumptions we made

- it is unbiased
- the variance
$$
\sigmaﾂｲ_{\bar{X}} = \frac{\sigmaﾂｲ}{n} \rightarrow 0 \qquad \text{ as } n \rightarrow \infty
$$


# In-class Exercises

**Exercise 1**

Suppose $X = 100$ with probability 0.8 and X = 600 with probability
0.2. Find the mean, variance and standard deviation of $X$ .

**Exercise 2**

Consider random samples of size 25 from the random variable $X$ that
has mean 100 and variance 400. Give the mean, variance and
standard deviation of the mean $\bar{X}$ .