---
title: "Exploratory Data Analysis"
subtitle: "Philosophy, Politics, and Economics"
author: "Miguel A. Arranz"
date: today
date-format: long
format: 
    html:
        code-fold: true
        code-summary: "Show the code"
---

# Data Preparation



First, let's load the required packages for our analysis:

```{r}
#| label: packages
#| warning: FALSE
library(foreign)      # For reading various data formats
library(dplyr)        # For data manipulation
library(tidyverse)    # For data science tools
library(ggplot2)      # For creating graphs
library(rmarkdown)    # For document formatting
library(kableExtra)   # For table formatting
library(modelsummary) # For statistical summaries
library(plotly)       # For interactive plots
```

Load and examine the dataset:


```{r}
#| label: load-data
library(here)

# This works regardless of which folder the file is in!
df <- read.csv(here("Data", "gapminder_full.csv"))
names(df)
```

Get a summary of the data:

```{r}
#| label: datasummary
#| warning: FALSE
datasummary_skim(data=df)
```

# Box Plots

A box plot (also known as a box-and-whisker plot) provides a visual summary of data distribution through five key statistics:

1. Minimum (within 1.5 × IQR)
2. First quartile (Q1, 25th percentile)
3. Median (Q2, 50th percentile)
4. Third quartile (Q3, 75th percentile)
5. Maximum (within 1.5 × IQR)

The interquartile range (IQR) is calculated as Q3 - Q1. Points beyond 1.5 × IQR from the quartiles are considered potential outliers and plotted individually.

## Life Expectancy Analysis

```{r}
#| label: ggboxplot
ggplot(df, aes(y=life_exp)) + 
  geom_boxplot() + 
  labs(title="Distribution of Life Expectancy",
       y="Life Expectancy (years)") -> p 
ggplotly(p)
```

```{r}
summary(df$life_exp)
```

Calculate the boundaries for outliers:

```{r}
q1 <- quantile(df$life_exp, 0.25)
q3 <- quantile(df$life_exp, 0.75)
iqr <- q3 - q1
upper <- q3 + 1.5*iqr
lower <- q1 - 1.5*iqr
```

|  Frontiers  |    |
| --- | --- |
| Lower | `r lower` | 
| Upper | `r upper` |

## GDP per Capita Analysis

```{r}
ggplot(df, aes(y=gdp_cap)) + 
  geom_boxplot() +
  labs(title="Distribution of GDP per Capita",
       y="GDP per Capita") -> p 
ggplotly(p)
```

```{r}
summary(df$gdp_cap)
```

Calculate the boundaries for outliers:

```{r}
q1 <- quantile(df$gdp_cap, 0.25)
q3 <- quantile(df$gdp_cap, 0.75)
iqr <- q3 - q1
upper <- q3 + 1.5*iqr
lower <- q1 - 1.5*iqr
```

|  Frontiers |    |
| --- | --- |
| Lower | `r lower` | 
| Upper | `r upper` |

# Logarithmic Transformation of GDP

Due to the skewed distribution of GDP per capita, we create a logarithmic transformation:

```{r}
df$log_gdp <- log(df$gdp_cap)
```

```{r}
ggplot(df, aes(y=log_gdp)) + 
  geom_boxplot() +
  labs(title="Distribution of Log GDP per Capita",
       y="Log GDP per Capita") -> p 
ggplotly(p)
```

```{r}
summary(df$log_gdp)
```

Calculate the boundaries for outliers:

```{r}
q1 <- quantile(df$log_gdp, 0.25)
q3 <- quantile(df$log_gdp, 0.75)
iqr <- q3 - q1
upper <- q3 + 1.5*iqr
lower <- q1 - 1.5*iqr
```

|  Frontiers |    |
| --- | --- |
| Lower | `r lower` | 
| Upper | `r upper` |

# Analysis by Continent

Compare life expectancy across continents:

```{r}
ggplot(df, aes(x=continent, y=log_gdp)) + 
  geom_boxplot() +
  labs(title="Log GDP per Capita by Continent",
       x="Continent",
       y="Log GDP per Capita") -> p 
ggplotly(p)
```

# Distribution Analysis

## Histograms

```{r}
#| warning: false
#| message: false
ggplot(df, aes(x=life_exp)) + 
  geom_histogram() +
  labs(title="Distribution of Life Expectancy",
       x="Life Expectancy (years)",
       y="Count")
```

With continental breakdown:

```{r}
#| warning: false
#| message: false
ggplot(df, aes(x=life_exp, fill=continent, color=continent)) + 
  geom_histogram() +
  labs(title="Distribution of Life Expectancy by Continent",
       x="Life Expectancy (years)",
       y="Count")
```

## Density Plots

A density plot shows the distribution of continuous data, where:
- The area under the curve equals 1
- The height at any point represents the relative likelihood of observations
- The shape indicates characteristics like skewness and modality

```{r}
ggplot(df, aes(x=life_exp)) + 
  geom_density() +
  labs(title="Density of Life Expectancy",
       x="Life Expectancy (years)",
       y="Density")
```

By continent:

```{r}
ggplot(df, aes(x=life_exp, color=continent)) + 
  geom_density() +
  labs(title="Density of Life Expectancy by Continent",
       x="Life Expectancy (years)",
       y="Density")
```

# Violin Plots

Violin plots combine box plots with kernel density estimates, showing:
- The median and quartiles (box plot elements)
- The full distribution shape (density estimate)
- Potential outliers
- Comparison across groups

```{r}
#| warning: false
#| message: false
#| output: false
library(ggpubr)

box <- ggplot(df, aes(x = "Box Plot", y = life_exp)) +
  geom_boxplot() +
  labs(y="Life Expectancy (years)")

violin <- ggplot(df, aes(x = "Violin Plot", y = life_exp)) +
  geom_violin(trim = FALSE, fill = "grey") +
  labs(y="Life Expectancy (years)")

ggarrange(box, violin, 
          ncol = 2, 
          common.legend = TRUE, 
          legend = "top",
          widths = c(2, 1), 
          heights = c(1, 1), 
          labels = c("A", "B"),
          font.label = list(size = 12, color = "black"),
          font.legend = list(size = 12, color = "black"),
          title = "Life Expectancy Distribution") -> p
```

```{r}
p
```

By continent with embedded box plots:

```{r}
ggplot(df, aes(x=continent, y=life_exp)) + 
  geom_violin() +
  geom_boxplot(width=0.01) +
  labs(title="Life Expectancy Distribution by Continent",
       x="Continent",
       y="Life Expectancy (years)") -> p
ggplotly(p)
```

# Normal Distribution Analysis

## QQ Plots

QQ (Quantile-Quantile) plots compare sample distributions to theoretical distributions:
- Points on the diagonal line suggest normal distribution
- Deviations indicate non-normality
- Useful for identifying outliers and distribution shape

GDP per capita:

```{r}
ggplot(df, aes(sample=gdp_cap)) + 
  stat_qq() + 
  stat_qq_line() +
  labs(title="Normal Q-Q Plot of GDP per Capita")
```

Log-transformed GDP per capita:

```{r}
ggplot(df, aes(sample=log_gdp)) + 
  stat_qq() + 
  stat_qq_line() +
  labs(title="Normal Q-Q Plot of Log GDP per Capita")
```

# Relationship Analysis

## Scatter Plots

Basic relationship between GDP and life expectancy:

```{r}
ggplot(df, aes(x=log_gdp, y=life_exp)) + 
  geom_point() +
  labs(title="Life Expectancy vs Log GDP per Capita",
       x="Log GDP per Capita",
       y="Life Expectancy (years)")
```

With continental breakdown:

```{r}
ggplot(df, aes(x=log_gdp, y=life_exp, color=continent)) + 
  geom_point() +
  labs(title="Life Expectancy vs Log GDP per Capita by Continent",
       x="Log GDP per Capita",
       y="Life Expectancy (years)") -> p
ggplotly(p)
```

Including population size:

```{r}
ggplot(df, aes(x=log_gdp, y=life_exp, 
               color=continent, 
               size=population)) + 
  geom_point() +
  labs(title="Life Expectancy vs Log GDP per Capita",
       subtitle="By Continent and Population Size",
       x="Log GDP per Capita",
       y="Life Expectancy (years)") -> p
ggplotly(p)
```
