---
title: "Introduction to R"
subtitle: "Philosophy, Politics, and Economics"
author: "Miguel A. Arranz"
date: today
date-format: long
format: 
    html:
        code-fold: true
        code-summary: "Show the code"
---

# Introduction

The objectives of this session are straightforward:

1. To read a data file in CSV format
2. To obtain some basic statistics
3. To create various types of graphs, such as histograms, boxplots, density plots, and violin plots
4. To subset a dataset
5. To visually compare different aspects of your dataset

To start, it's important to create a file to keep a record of all the commands you'll execute in R. This way, you can easily replicate your results later on.

To create a file in RStudio:

1. Go to the File menu and select "New File," then "R Script"
2. Save your file (File → Save As) in the same folder as your data file
3. Set the working directory (Session → Set Working Directory → File Location)

As you progress through the course, you'll learn how to use RMarkdown to unlock the full potential of R, but for now, let's keep things simple and focus on the statistical concepts first.

For a more spacious and uncluttered screen, click the "Show in New Window" icon located below the file name. This will give you two windows: one with your script file and another with the three remaining panels of RStudio.

Important tips:
- To run commands: Write them in the file, select the desired lines, and click the "Run" icon
- To add comments: Start the line with the "#" symbol
- Save your file frequently to prevent loss of work

# Reading Data

R is a powerful programming language primarily used for data science. Unlike Python, which has broader applications, R specializes in statistical analysis and data manipulation.

R can import data from many different file formats, which is a significant advantage over other statistical programs. To import different file formats, you'll need to use specialized packages. One such package is "foreign," which comes pre-installed with R.

Let's read our dataset:

```{r}
library(here)
df <- read.csv(here("Data", "gapminder_full.csv"))
```

The command `read.csv()` reads CSV files into R. We store the data in a data frame called "df". For help with any command, use `help(command_name)` or `?command_name` in the console.

To explore the loaded data:
- Use `names(df)` to see column names
- Use `str(df)` to examine the data structure

# Installing Packages

To enhance R's capabilities, we'll need additional packages. Install packages from the console using:

```r
install.packages("modelsummary")
install.packages("vioplot")
```

After installation, load the packages into your session:

```{r}
#| warning: false
library(modelsummary)
library(vioplot)
```

Note: It's best to install packages from the console rather than the script file, as it might require repository selection.

# Getting the Basic Statistics

Let's start with basic statistical analysis. To analyze the variable "life_exp":

```{r}
#| output: true
mean(df$life_exp)
median(df$life_exp)
quantile(df$life_exp)
var(df$life_exp)
sd(df$life_exp)  # Fixed: Removed nested var() function
```

To store results for later use, assign them to variables:

```{r}
xbar <- mean(df$life_exp)
```

## Exercise

Can you explain why `s` and `sigma` are equal in the following code?

```{r}
#| output: true
s2 <- var(df$life_exp)
s <- sqrt(s2)
sigma <- sd(df$life_exp)
print(s)
print(sigma)
```

The values are equal because the standard deviation (`sd`) is the square root of the variance. We can calculate it either way:
- Calculate variance and take its square root
- Use the `sd()` function directly

## Objects with Multiple Elements

Some statistical functions return multiple values:

```{r}
#| output: true
S <- summary(df$life_exp)
Q <- quantile(df$life_exp)
print(names(S))
print(names(Q))
print(S)
print(Q)
```

To access specific values, use either names or positions:

```{r}
#| output: true
m1 <- S["Median"]
m2 <- S[3]
print(m1)
print(m2)

m1a <- Q["50%"]
m2a <- Q[3]
print(m1a)
print(m2a)
```

# Summary of the Dataframe

For a complete overview of the data, use either base R's `summary()` or the more sophisticated `datasummary_skim()`:

```{r}
#| output: true
summary(df)
```

```{r}
#| output: true
datasummary_skim(df)
```

# Graphics

Let's create various visualizations of our data.

## Histogram

Basic histogram:

```{r}
#| output: true
hist(df$life_exp, main="Life Expectancy")  # Fixed spelling of "Expectancy"
```

## Density Plots

```{r}
#| output: true
plot(density(df$life_exp), main="Life Expectancy")  # Fixed spelling
```

## Boxplot

```{r}
#| output: true
boxplot(df$gdp_cap, main="GDP per Capita")
```

## Boxplots with Factors

Compare life expectancy across continents:

```{r}
#| output: true
boxplot(df$life_exp ~ df$continent, main="Life Expectancy")  # Fixed spelling
```

## Violin Plots

```{r}
#| output: true
vioplot(df$life_exp, main="Life Expectancy")  # Fixed spelling
```

```{r}
#| output: true
vioplot(df$life_exp ~ df$continent, main="Life Expectancy")  # Fixed spelling
```

# Subsetting the Dataset

To work with specific parts of your data, you can create subsets:

```{r}
df_Asia1 <- df[df$continent=="Asia", ]
df_Asia2 <- subset(df, continent=="Asia")
```

Compare the results:

```{r}
datasummary_skim(df_Asia1)
```

```{r}
datasummary_skim(df_Asia2)
```

Check the number of observations:

```{r}
print(length(df_Asia1$life_exp))
```

View the distribution across continents:

```{r}
table(df$continent)
```
