---
title: "Inference on the Mean"
subtitle: "Philosophy, Politics, and Economics"
author: "Miguel A. Arranz"
date: today
date-format: long
format: 
    html:
        code-fold: true
        code-summary: "Show the code"
---


# Packages

```{r warning=FALSE, message=FALSE}
library(foreign)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(rmarkdown)
library(kableExtra)
library(modelsummary)
library(plotly)
library(haven)
library(ggplot2)
library(ggthemes)
library(extrafont)
```


# Review

## Central Limit Theorem

$$
\bar{X} \overset{a}{\sim} N \left(\mu, \frac{\sigma^2}{n} \right)
$$

Transform

$$
\frac{\bar{X} - \mu}{\sigma/\sqrt{n}} \sim N(0,1)
$$

$N(0,1)$ is the **standard Normal distribution**

Unknow $\sigma$


$$
\frac{\bar{X} - \mu}{s/\sqrt{n}} \sim t_\nu
$$

This what we call the t-stat

$$
\frac{\hat{\theta} - \theta}{se(\hat{\theta})} \sim t_\nu
$$

However, if $X$ does not follow Normal distribution, it is better to go for the **asymptotic distribution**


$$
\frac{\bar{X} - \mu}{s/\sqrt{n}} \sim N(0,1)
$$

and then


$$
\frac{\bar{X} - \mu}{s/\sqrt{n}} \sim N(0,1)
$$


## Graphical Analysis

$$
\mu = 1000, \; \sigma=250
$$

$$
Y = \bar{X}
$$

$$
\Pr( 500 \leq Y \leq 1000) = \int_{y=500}^{y=1000} \phi(y) dy = F_Y(1000) - F_Y(500)
$$

$$\Pr(Y \leq 500) = F_Y(500)$$

$$\Pr(Y > 500) = 1 - \Pr(Y \leq 500) = 1 - F_Y(500)$$


```{r}
#| warning: false
mean <- 1000
std <- 250
x <- seq(mean - 3 * std, mean + 3 * std, length.out = 1000)
y <- dnorm(x, mean, std)

data <- data.frame(x = x, y = y)

px <- seq(0, 500, by = 10)
py <- dnorm(px, mean, std)

fill_data <- data.frame(x = px, y = py)

ggplot() +
  geom_line(data = data, aes(x = x, y = y), color = "blue", size = 1.2) +
  geom_ribbon(data = fill_data, aes(x = x, ymin = 0, ymax = y), fill = "red", alpha = 0.5) +
  labs(title = "Normal Distribution", x = "Value", y = "Density")

```




```{r}
#| warning: false
library(ggplot2)

mean <- 1000
std <- 250
x <- seq(mean - 3 * std, mean + 3 * std, length.out = 1000)
y <- dnorm(x, mean, std)

data <- data.frame(x = x, y = y)

px <- seq(1500, 2000, by = 10)
py <- dnorm(px, mean, std)

fill_data <- data.frame(x = px, y = py)

ggplot() +
  geom_line(data = data, aes(x = x, y = y), color = "blue", size = 1.2) +
  geom_ribbon(data = fill_data, aes(x = x, ymin = 0, ymax = y), fill = "red", alpha = 0.5) +
  labs(title = "Normal Distribution", x = "Value", y = "Density")

```


```{r}
#| warning: false

mean <- 1000
std <- 250
x <- seq(mean - 3 * std, mean + 3 * std, length.out = 1000)
y <- dnorm(x, mean, std)

data <- data.frame(x = x, y = y)

px1 <- seq(0, 500, by = 10)
py1 <- dnorm(px1, mean, std)

fill_data1 <- data.frame(x = px1, y = py1)

px2 <- seq(1500, 2000, by = 10)
py2 <- dnorm(px2, mean, std)

fill_data2 <- data.frame(x = px2, y = py2)

ggplot() +
  geom_line(data = data, aes(x = x, y = y), color = "blue", size = 1.2) +
  geom_ribbon(data = fill_data1, aes(x = x, ymin = 0, ymax = y), fill = "red", alpha = 0.5) +
  geom_ribbon(data = fill_data2, aes(x = x, ymin = 0, ymax = y), fill = "red", alpha = 0.5) +
  labs(title = "Normal Distribution", x = "Value", y = "Density")



```

# Student's t and Normal

```{r}
#| warning: false


x <- seq(-4, 4, length=100)
hx <- dnorm(x)
labels <- c("t with df = 5", "standard normal")

df <- data.frame(x = rep(x, 2), 
                 density = c(hx, dt(x, 4)), 
                 distribution = rep(labels, each = length(x)))

ggplot(df, aes(x = x, y = density, color = distribution, linetype = distribution)) +
  geom_line(size = 1) +
  labs(x = "x value", y = "Density", title = "t(4) and standard normal") +
  scale_linetype_manual(values = c(2, 1)) +
  scale_color_manual(values = c("red", "blue")) +
  theme_minimal() +
  theme(legend.position = "top", 
        legend.title = element_text(face = "bold"),
        legend.text = element_text(size = 10))


```

```{r}
#| warning: false


x <- seq(-4, 4, length=100)
hx <- dnorm(x)
labels <- c("t with df = 30", "standard normal")

df <- data.frame(x = rep(x, 2), 
                 density = c(hx, dt(x, 30)), 
                 distribution = rep(labels, each = length(x)))

ggplot(df, aes(x = x, y = density, color = distribution, linetype = distribution)) +
  geom_line(size = 1) +
  labs(x = "x value", y = "Density", title = "t(30) and standard normal") +
  scale_linetype_manual(values = c(2, 1)) +
  scale_color_manual(values = c("red", "blue")) +
  theme_minimal() +
  theme(legend.position = "top", 
        legend.title = element_text(face = "bold"),
        legend.text = element_text(size = 10))

```



# Hypothesis Testing

$$
H_0 \colon \: \mu=\mu_0 \quad vs. \quad H_1 \colon \; \mu \neq \mu_0
$$


$$
H_0 \colon \: \mu=\mu_0 \quad vs. \quad H_1 \colon \; \mu > \mu_0
$$


$$
H_0 \colon \: \mu=\mu_0 \quad vs. \quad H_1 \colon \; \mu < \mu_0
$$


## Case 1


$$
H_0 \colon \: \mu=\mu_0 \quad vs. \quad H_1 \colon \; \mu \neq \mu_0
$$

Assume that $\mu_0=1000$. 

Your estimator is $\bar{x}=300$. The estimated value for the the variance of $\bar{x}$ is 250


```{r}
#| warning: false

mean <- 1000
std <- 250
x <- seq(mean - 3 * std, mean + 3 * std, length.out = 1000)
y <- dnorm(x, mean, std)

data <- data.frame(x = x, y = y)

px1 <- seq(0, 300, by = 10)
py1 <- dnorm(px1, mean, std)

fill_data1 <- data.frame(x = px1, y = py1)

px2 <- seq(1700, 2000, by = 10)
py2 <- dnorm(px2, mean, std)

fill_data2 <- data.frame(x = px2, y = py2)

ggplot() +
  geom_line(data = data, aes(x = x, y = y), color = "blue", size = 1.2) +
  geom_ribbon(data = fill_data1, aes(x = x, ymin = 0, ymax = y), fill = "red", alpha = 0.5) +
  geom_ribbon(data = fill_data2, aes(x = x, ymin = 0, ymax = y), fill = "red", alpha = 0.5) +
  labs(title = "Normal Distribution", x = "Value", y = "Density")



```


Your estimator is $\bar{x}=1200$. The estimated value for the the variance of $\bar{x}$ is 250


```{r}
#| warning: false

mean <- 1000
std <- 250
x <- seq(mean - 3 * std, mean + 3 * std, length.out = 1000)
y <- dnorm(x, mean, std)

data <- data.frame(x = x, y = y)

px1 <- seq(0, 800, by = 10)
py1 <- dnorm(px1, mean, std)

fill_data1 <- data.frame(x = px1, y = py1)

px2 <- seq(1200, 2000, by = 10)
py2 <- dnorm(px2, mean, std)

fill_data2 <- data.frame(x = px2, y = py2)

ggplot() +
  geom_line(data = data, aes(x = x, y = y), color = "blue", size = 1.2) +
  geom_ribbon(data = fill_data1, aes(x = x, ymin = 0, ymax = y), fill = "red", alpha = 0.5) +
  geom_ribbon(data = fill_data2, aes(x = x, ymin = 0, ymax = y), fill = "red", alpha = 0.5) +
  labs(title = "Normal Distribution", x = "Value", y = "Density")



```

## Case 2


$$
H_0 \colon \: \mu=\mu_0 \quad vs. \quad H_1 \colon \; \mu > \mu_0
$$




Your estimator is $\bar{x}=1200$. The estimated value for the the variance of $\bar{x}$ is 250



```{r}
#| warning: false
library(ggplot2)

mean <- 1000
std <- 250
x <- seq(mean - 3 * std, mean + 3 * std, length.out = 1000)
y <- dnorm(x, mean, std)

data <- data.frame(x = x, y = y)

px <- seq(1200, 2000, by = 10)
py <- dnorm(px, mean, std)

fill_data <- data.frame(x = px, y = py)

ggplot() +
  geom_line(data = data, aes(x = x, y = y), color = "blue", size = 1.2) +
  geom_ribbon(data = fill_data, aes(x = x, ymin = 0, ymax = y), fill = "red", alpha = 0.5) +
  labs(title = "Normal Distribution", x = "Value", y = "Density")

```

What if $\bar{x}=800$?


```{r}
#| warning: false
library(ggplot2)

mean <- 1000
std <- 250
x <- seq(mean - 3 * std, mean + 3 * std, length.out = 1000)
y <- dnorm(x, mean, std)

data <- data.frame(x = x, y = y)

px <- seq(800, 2000, by = 10)
py <- dnorm(px, mean, std)

fill_data <- data.frame(x = px, y = py)

ggplot() +
  geom_line(data = data, aes(x = x, y = y), color = "blue", size = 1.2) +
  geom_ribbon(data = fill_data, aes(x = x, ymin = 0, ymax = y), fill = "red", alpha = 0.5) +
  labs(title = "Normal Distribution", x = "Value", y = "Density")

```



What if $\bar{x} = 1500$?

```{r}
#| warning: false
library(ggplot2)

mean <- 1000
std <- 250
x <- seq(mean - 3 * std, mean + 3 * std, length.out = 1000)
y <- dnorm(x, mean, std)

data <- data.frame(x = x, y = y)

px <- seq(1500, 2000, by = 10)
py <- dnorm(px, mean, std)

fill_data <- data.frame(x = px, y = py)

ggplot() +
  geom_line(data = data, aes(x = x, y = y), color = "blue", size = 1.2) +
  geom_ribbon(data = fill_data, aes(x = x, ymin = 0, ymax = y), fill = "red", alpha = 0.5) +
  labs(title = "Normal Distribution", x = "Value", y = "Density")

```


## Case 3


$$
H_0 \colon \: \mu=\mu_0 \quad vs. \quad H_1 \colon \; \mu < \mu_0
$$


$\bar{x}$ = 500

```{r}
#| warning: false
mean <- 1000
std <- 250
x <- seq(mean - 3 * std, mean + 3 * std, length.out = 1000)
y <- dnorm(x, mean, std)

data <- data.frame(x = x, y = y)

px <- seq(0, 500, by = 10)
py <- dnorm(px, mean, std)

fill_data <- data.frame(x = px, y = py)

ggplot() +
  geom_line(data = data, aes(x = x, y = y), color = "blue", size = 1.2) +
  geom_ribbon(data = fill_data, aes(x = x, ymin = 0, ymax = y), fill = "red", alpha = 0.5) +
  labs(title = "Normal Distribution", x = "Value", y = "Density")

```

$\bar{x} = 300$


```{r}
#| warning: false
mean <- 1000
std <- 250
x <- seq(mean - 3 * std, mean + 3 * std, length.out = 1000)
y <- dnorm(x, mean, std)

data <- data.frame(x = x, y = y)

px <- seq(0, 300, by = 10)
py <- dnorm(px, mean, std)

fill_data <- data.frame(x = px, y = py)

ggplot() +
  geom_line(data = data, aes(x = x, y = y), color = "blue", size = 1.2) +
  geom_ribbon(data = fill_data, aes(x = x, ymin = 0, ymax = y), fill = "red", alpha = 0.5) +
  labs(title = "Normal Distribution", x = "Value", y = "Density")

```

$bar{x} = 1300$ 


```{r}
#| warning: false
mean <- 1000
std <- 250
x <- seq(mean - 3 * std, mean + 3 * std, length.out = 1000)
y <- dnorm(x, mean, std)

data <- data.frame(x = x, y = y)

px <- seq(0, 1300, by = 10)
py <- dnorm(px, mean, std)

fill_data <- data.frame(x = px, y = py)

ggplot() +
  geom_line(data = data, aes(x = x, y = y), color = "blue", size = 1.2) +
  geom_ribbon(data = fill_data, aes(x = x, ymin = 0, ymax = y), fill = "red", alpha = 0.5) +
  labs(title = "Normal Distribution", x = "Value", y = "Density")

```

## p-value

The p-value is the probability of observing a test statistic as extreme as, or more extreme than, the one obtained from the sample, assuming that the null hypothesis ($H_0$) is true.

In simpler terms, the p-value measures the strength of evidence against $H_0$

​
A small p-value (typically < 0.05) suggests that the observed data is unlikely under $H_0$, leading to reject $H_0$
​
 
A large p-value (typically > 0.05) indicates that the observed data is consistent with $H_0$, so we do not reject it.



# Confidence Interval


$$
\frac{\bar{X} - \mu}{s/\sqrt{n}} \sim N(0,1)
$$

$$
\begin{aligned}
\Pr \left( \left|  \frac{\bar{X} - \mu}{s/\sqrt{n}} \right| \leq z_{\alpha/2} \right) = 1 - \alpha\\
\Pr \left( \left|  \frac{\bar{X} - \mu}{s/\sqrt{n}} \right| \leq z_{\alpha/2} \right) =
\Pr \left( -z_{\alpha/2} \leq \frac{\bar{X} - \mu}{s/\sqrt{n}}  \leq z_{\alpha/2} \right) 
\end{aligned}
$$

$$
-z_{\alpha/2} \leq \frac{\bar{x} - \mu}{s/\sqrt{n}}  \leq z_{\alpha/2} \quad \rightarrow 
-z_{\alpha/2} \times \frac{s}{\sqrt{n}} \leq \bar{x} - \mu \leq z_{\alpha/2} \times \frac{s}{\sqrt{n}}
$$

$$
-\bar{x} -z_{\alpha/2} \times \frac{s}{\sqrt{n}} \leq - \mu \leq -\bar{x} + z_{\alpha/2} \times \frac{s}{\sqrt{n}}
$$

$$
\bar{x} -z_{\alpha/2} \times \frac{s}{\sqrt{n}} \leq  \mu \leq \bar{x} + z_{\alpha/2} \times \frac{s}{\sqrt{n}}
$$

Confidence Interval

$$
\left ({\bar{x}} \pm z_{/\alpha/2} \times \frac{s}{\sqrt{n}} \right)
$$

In general
$$
\alpha=0.05 \qquad z_{\alpha/2} = 1.96
$$

$$
\hat{\theta} \pm 2 \times se(\hat{\theta})
$$
